---
title: "Affective visualization graphs"
author: "Callie Baker"
date: "2026-06-01"
format:
  pdf:
    toc: true # use this to display a table of contents
execute:
  message: false # use this to make sure messages don't show up
  warning: false # use this to make sure warnings don't show upl
---

# Setting up

```{r}
#| label: loading-packages

# read in packages
library(tidyverse) # general use
library(here) # file organization -- dont need unless sort into folders
library(janitor) # cleaning data frames
library(DHARMa) # for diagnostics
library(ggeffects) # for ggpredict
library(flextable) # for flextable
library(gtsummary)

```

```{r}
#| label: reading-data

# read in personal data to object from csv
my_data <- read_csv("envs193ds-personal-data.csv")

```


# Framing my question
My central question is: How does sleep duration the night prior influence the probability of me going on a run the next day? 

Variables

- Predictor: sleep duration the previous night (hours) 
- Response: probability of going on a run (1/0) (1 = yes; 0 = no run) 


# Cleaning and wrangling data

```{r}
#| label: wrangling-data

# cleaned column names
my_data <- my_data |> 
  clean_names()

# created new column: 1 if workout was a run, 0 if not
my_data <- my_data |>
  mutate(is_run = case_when(
    workout_type == "Run" ~ 1, # run = 1
    .default = 0))             # all other workout types = 0

```


# Fitting the model

```{r}
#| label: model-glm

# fitted generalized linear model for running
# response: whether or not workout was a run (binary: 1/0)
# predictor: sleep the previous night (hours)
run_mod <- glm(
  # formula: binary run outcome ~ sleep hours
  is_run ~ sleep_hours,
  # data: my_data data frame
  data = my_data,
  # binomial error distribution for binary response
  family = binomial)

```


# Checking diagnostics
```{r}
#| label: diagnostics
#| fig-height: 6
#| fig-width: 8

# simulated residuals from GLM for diagnostic plots
plot(simulateResiduals(run_mod))

```

**Info from left plot (QQ plot of residuals)**:

- KS test: do my residuals follow a uniform distribution? → generally yes (CHECK)
- Dispersion test: are your residuals more dispersed than expected from a uniform distribution? → no
- Outlier test: are there outliers in your data set that could be influencing model coefficients or predictions? → no

**Info from right plot (residuals vs predicted values)**:

- By default, R generates the reference lines for quartiles: 25, 50, 75
- Not that informative but look at patterns within residuals across the range → DHARMa diagnostics flag issues in 25th quartile due to small sample size, not because the model is wrong. 


# Looking at the model coefficients
```{r}
#| label: model summary

# displayed model summary
summary(run_mod)

```


## Calculate inverse logit
```{r}
#| label: coefficients-b0

# annotate 
plogis(-8.7011)

```
**Note**: 
- inverse logit is 0.00016
- The probability of going on a run if sleep duration is 0 hours is 0.00016??


## Calculate odds ratio 

```{r}
#| label: coefficients-b1

# annotate 
gtsummary::tbl_regression(run_mod, exponentiate = TRUE)

```
**Note**: 
- odds ratio is 2.97
- 2.97 > 1 so the probability of going on a run increases by a factor of 2.97 with every 1 hour increase in sleep

**Summary**: 
- For every 1 hour increase in sleep, the the odds of going on a run increase by a factor of 2.97 (95% CI: [1.37, 9.40], p = 0.021, ⍺ = 0.05). ?? NOT SIGNIF THO


## Other way to calculate -- remove?
```{r}
#| label: coefficients

# exponentiated coefficients to get odds ratios
# e.g. exp(coef) = odds ratio for 1 hr increase in sleep
exp(coef(run_mod))

```


# Looking at the model predictions
- using probabilities to determine chance of getting certain outcome
- line represents the probability of 1 (going on a run) actually occurring

```{r}
#| label: model-predictions

# what is the probabilty of going on a run after sleeping for 10 hours?
ggpredict(run_mod,
          terms = "sleep_hours [10]")

# what is the probabilty of going on a run after sleeping for 4 hours?
ggpredict(run_mod,
          terms = "sleep_hours [4]")

```

**Note**:

- If I slept for 10 hours, the probability of going on a run the next day is 0.90 (95% CI [0.39, 0.99]).
- If I slept for 4 hours, the probability of going on a run the next day is 0.01 (95% CI [0.00, 0.25]).


# Visualizing model predictions

```{r}
#| label: visualizing-model
#| fig-height: 6
#| fig-width: 8

# generated model predictions across range of sleep values in data set
mod_preds <- ggpredict(run_mod,
                       terms = "sleep_hours [5:10 by = 1]")

# base layer: ggplot with raw data
# data frame: my_data
ggplot(my_data,
       aes(x = sleep_hours,
           y = is_run)) +
  # first layer: raw observations
  geom_point(size = 3,
             alpha = 0.4) +
  # second layer: confidence interval ribbon from model predictions
  geom_ribbon(data = mod_preds,
              aes(x = x,
                  y = predicted,
                  ymin = conf.low,
                  ymax = conf.high),
                  alpha = 0.4) +
  # third layer: logistic curve from model predictions
  geom_line(data = mod_preds,
            aes(x = x,
                y = predicted),
            # custom color
            color = "navyblue") +
  # constrained y-axis to 0 and 1
  scale_y_continuous(limits = c(0, 1),
                     breaks = c(0, 1)) +
  # relabeled axes and added title
  labs(title = "More sleep is associated with higher probability of running the next day",
       x     = "Sleep the previous night (hours)",
       y     = "Probability of going on a run (1 = yes, 0 = no)") +
  # changed theme from default
  theme_minimal() +
  # bolded title
  theme(plot.title = element_text(face = "bold"))

```


# Describing the model

```{r}
#| label: describing-model

# displaying model in table
as_flextable(run_mod)

```



